Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques

### 2019-11-28 paper/paperV3.tex
Final V3

Scenarios generated by \textit{simScene} should cover most real-life situations while remaining perceptually plausible. To guarantee the realism of created sound scenes, the information needed for \textit{simScene} generation is based on a corpus of 74 recordings from the GRAFIC project~\cite{aumond2017}. These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, and range from 55~s to 4.5~mn in duration. In~\cite{gloaguen2017}, the extracts are manually annotated and classified in terms of ambiance (\textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}). The available annotations include:

@@

The 75 simulated scenes of the listening test corpus are generated using \textit{simScene} in the \textit{generate} mode with the extracted distributions as input. To improve the diversity of scenarios considered in the listening test, 200 sound scenes are first simulated, equally distributed among the five ambiances (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}). A common duration of 45 seconds is chosen due to the limited duration of listening tests because of the fatigue. Though many studies dealing with the environmental acoustic quality are associated with acoustic measurements ranging from a few seconds~\cite{paulsen1997, brambilla2006} to fifteen minutes~\cite{kuwano1997, decoensel2007} and even to eighty minutes~\cite{namba1988}, stimuli between 30 seconds and 1 minute are often preferred for laboratory tests. To avoid simulated scenes to be too similar in the listening experiment corpus, a selection of 75 simulated scenes is carried out from the initial 200 scenes, in order to cover a large diversity of presence for each type of sources (ranging from no presence to 100\% of presence) without redundancy. Playback sound levels are realistically drawn for each scene, and range from 46.6~dB SPL to 77.1~dB SPL over the 75 simulated scenes.

In order to validate the generation procedure of new scenarios, the listening test corpus is extended with replications of recorded sound scenes. Reference recordings are obtained from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations (noted P1-19) with diverse environments. The path undertaken during these soundwalks is shown in Figure~\ref{fig:path}, with sound levels in dB SPL interpolated from measurements at the 19 locations. For each of the 19 corresponding recordings, 45 seconds of audio in a single channel are extracted. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The manual annotations available in~\cite{gloaguen2017} of background and event information are then used to replicate the sound scenes using \textit{simScene} in the \textit{replicate} mode. The 19 resulting sound scenes are thus simulated, but their scenarios follow those of the reference recordings to the extent of annotation precision. Furthermore, the original 45~s recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization. The 6 recorded and 19 replicated scenes are scaled using the relation found in Section~\ref{sec:exp_equi} so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL over the 19 locations.

@@

During the test the scenes are played at a given sound level as discussed in Section~\ref{sec:exp_stim}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used by all participants. The calibration of the headphones was carried out in a free field situation (pink noise through a Genelec 1031A loudspeaker in front of the head) and consisted in characterizing the relationship between voltage at the headphone terminal and the corresponding binaural sound pressure. To do so, the following procedure was conducted:
\begin{enumerate}
\item A pink noise generator is set at an arbitrary level and its output voltage is measured.
\item Small DPA 4060 microphones are set at the entrance of the ear canals of a human participant~\cite{moller1992}. The generator is used as input of the headphones, placed over the head of the participant. The voltage at the output of the binaural microphones is measured.
\item The headphones are removed, and the generator is used as input of Genelec 1031A loudspeaker placed at a distance of 1 meter from the participant's head. The loudspeaker's amplification is tuned until the same voltage is outputted by the binaural microphones.
\item The head is replaced by a sound level meter to measure the sound level in dB of the loudspeaker. This corresponds to the binaural sound level produced by the headphones for the considered output voltage of the pink noise generator.
\end{enumerate}
By repeating this procedure for different settings of the generator level, the relation between the logarithm of the generator output voltage and headphones playback sound level in dB is approximated as a linear function. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener, when using the calibrated headphones and a specific sound card configuration for which the relation between the numeric level and output voltage is known.

@@

\begin{table*}[t!]
\centering
\caption{Mean differences of perceptual assessments (resp. pleasantness, liveliness, overall loudness, interest, calmness, time of presence of traffic, voices and birds) between original and replicated sound scenes. Significant differences as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
\label{tab:ogrep}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P1 & 0.43 & \textbf{-1.65} & \textbf{-1.04} & 0.43 & 0.13 & 0.39 & \textbf{-2.09} & 0.61 \\
	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 1.04 & \textbf{-4} & 0.22 \\
	P4 & 0.91 & 0 & \textbf{-1.83} & 0.48 & \textbf{1.30} & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-0.91} & 0.09 & \textbf{-1.43} \\
	P15 & \textbf{-1.35} & 0.52 & 0.52 & \textbf{-1.17} & 0.09 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
	P18 & \textbf{1.13} & -0.30 & \textbf{-1.17} & -0.43 & \textbf{1.39} & \textbf{-1.83} & \textbf{0.83} & \textbf{1.30} \\ \hline
\end{tabular}
%}
\end{table*}

Participants are asked to evaluate sound scenes using 8 perceptual attributes represented by 11-point scales (0-10). These attributes are presented in French and translated in this report using standard terminology. The first 5 semantic differential scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, tranquil} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. Additionally, to assess the perceived source activity 3 questions are presented to the participants and evaluated on the same 11-point scales:
\begin{itemize}
\item Time of presence of traffic, voice and bird sources: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

@@

The statistic analyses carried out in this paper (statistic comparison tests, principal component analysis) are done with Matlab. The effect of the \textit{simScene} generation procedure on perception is first investigated. To do so, perceptual responses are compared for the 6 recorded scenes and 6 corresponding replicated scenes, which share common scene compositions. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the first five scales, all mean differences are lower than 2 points on the 11-point discrete scale. Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Though, overall no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.

@@

First, some indicators can be computed directly from the mixed audio. Typically this includes indicators derived from sound level measurements used in monitoring applications. For this study the following variables are considered, and computed with a time frame of 1~s using the Matlab ITA-toolbox~\cite{itatoolbox2017}:
\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is often associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\item $LA_{10}-LA_{90}$: Emergence indicator included in the pleasantness model presented in~\cite{ricciardi2014}.
\end{itemize}
The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of perceived source activity. Its expression is:
\begin{equation}
TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f_1=31.5Hz}^{f_1=16kHz}\lvert\frac{d^2L(f_1, t)}{df_1dt}\rvert}
\end{equation}
where $L(f, t)$ is the third-octave spectrum of the signal. This indicator represents the variations in both the time and frequency dimensions to highlight sources of interest. For example, bird activity is characterized by narrow-band energy with fast paced variations in time, which translates into high $TFSD$ values in the corresponding frequency range. The $TFSD$ is computed for the 4kHz band and 125ms measurements ($TFSD_{4kHz(1/8s)}$), and for the 500Hz band with 1s measurements ($TFSD_{500Hz, 1s}$) as estimates of the perceived activity of birds and voices respectively.

@@

\begin{equation}\label{eq:ts}
\hat T_s(\alpha, \beta) = \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta L_{s}(t, f)\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}>\beta \right]\\
\end{equation}
where $s$ denotes the sound source, $N_t$ is the number of considered time frames, $N_f$ is the number of frequency bands. This indicator evaluates the of percentage of presence of a given source with regard to other sources. First, for each time frame, frequency bands for which the source is emergent, that is the difference between its sound level and that of all other sources $\Delta L_{s}(t, f)$ is greater than the $\alpha$ threshold value, are isolated. Then, the source is considered present in this time frame if the emergence for these bands is on average greater than the $\beta$ threshold. A time of presence estimation is obtained for each source by averaging the resulting binary presence values along the $N_t$ time frames composing the scene. The optimal threshold values are found via grid search so that the time of presence estimation $\hat T_s(\alpha_{opt}, \beta_{opt})$ is best correlated with the average perceived time of presence $T_{s, p}$ as obtained by human evaluation:
\begin{equation}\label{eq:ts_opt}
\alpha_{opt},\beta_{opt} = \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, \hat T_s(\alpha, \beta)\right)
\end{equation}
where $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $r$ denotes the Pearson correlation coefficient and $T_{s, p}$ corresponds to the perceived time of presence assessments averaged per scene. Both thresholds are optimized once in this study using data from the listening test in Section~\ref{sec:exp}, and equal $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$, though several other pairs of values yielded close performance as per the optimization metric defined in eq.\ref{eq:ts}.

@@

\begin{table*}[ht!]
\centering
%\resizebox{\textwidth}{!}{
\caption{Pearson's correlation coefficients between physical and perceptual (resp. pleasantness, liveliness, overall loudness, interest, calmness, time of presence of traffic, voices and birds) indicators (n = 92).}
\label{tab:physc}
\begin{threeparttable}[t]
\begin{tabular}{ c | c c c c c | c c c c }
\hline
	 & P & L & OL & I & C & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	$LA_{eq}$ & -0.86** & 0.68** & 0.92** & -0.37** & -0.88** & 0.66** & 0.07 & -0.41** \\
	$LA_{50}$ & -0.84** & 0.67** & 0.91** & -0.33** & -0.87** & 0.63** & 0.06 & -0.35** \\
	$L_{eq}$ & -0.88** & 0.67** & 0.91** & -0.44** & -0.88** & 0.71** & 0.06 & -0.46** \\
	$L_{10}$ & -0.87** & 0.65** & 0.90** & -0.44** & -0.86** & 0.71** & 0.06 & -0.47** \\
	$L_{50}$ & -0.89** & 0.65** & 0.92** & -0.43** & -0.89** & 0.71** & 0.03 & -0.44** \\
	$L_{90}$ & -0.86** & 0.68** & 0.92** & -0.39** & -0.89** & 0.67** & 0.07 & -0.40** \\
	$L_{50, 1kHz}$ & -0.88** & 0.69** & 0.92** & -0.42** & -0.89** & 0.73** & 0.08 & -0.50** \\
	$L_{10}-L_{90}$ & 0.13 & -0.18 & -0.24* & -0.06 & -0.22* & -0.01 & -0.01 & -0.09 \\ \hline
	$TFSD_{500Hz, 1s}$ & 0.07 & 0.41** & 0.11 & 0.28** & -0.15 & -0.39** & 0.74** & -0.17 \\
	$TFSD_{4kHz, 1/8s}$ & 0.52** & -0.43** & -0.49** & 0.41** & 0.52** & -0.54** & -0.18 & 0.63** \\ \hline
	$L_{eq, T}$ & -0.58** & 0.20 & 0.46** & -0.46** & -0.42** & 0.71** & -0.16 & -0.36** \\
	$L_{eq, V}$ & -0.17 & 0.50** & 0.31** & 0.08 & -0.37** & -0.04 & 0.71** & -0.40** \\
	$L_{eq, B}$ & 0.27* & -0.04 & -0.11 & 0.35** & 0.18 & -0.24* & -0.04 & 0.71** \\ \hline
	$\Delta L_T$ & -0.45** & -0.11 & 0.26* & -0.59** & -0.22* & 0.66** & -0.51** & -0.26* \\
	$\Delta L_V$ & 0.04 & 0.50** & 0.17 & 0.35** & -0.20 & -0.38** & 0.59** & -0.01 \\
	$\Delta L_B$ & 0.21* & -0.25* & -0.26* & 0.08 & 0.25* & -0.25* & -0.10 & -0.03 \\ \hline
	$\hat T_T(\alpha_{opt}, \beta_{opt})$ & -0.53** & -0.05 & 0.35** & -0.57** & -0.29** & 0.81** & -0.39** & -0.37** \\
	$\hat T_V(\alpha_{opt}, \beta_{opt})$ & 0.12 & 0.44** & 0.05 & 0.35** & -0.11 & -0.39** & 0.81** & -0.16 \\
	$\hat T_B(\alpha_{opt}, \beta_{opt})$ & 0.56** & -0.30** & -0.46** & 0.55** & 0.51** & -0.57** & -0.08 & 0.91** \\ \hline
\end{tabular}
\begin{tablenotes}
\item *: p<0.05, **: p<0.01
\end{tablenotes}
%}
\end{threeparttable}
\end{table*}

>>>>

Scenarios generated by \textit{simScene} should cover most real-life situations while remaining perceptually plausible. To guarantee the realism of created sound scenes, the information needed for \textit{simScene} generation is based on a corpus of 74 recordings from the GRAFIC project~\cite{aumond2017}. These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, as seen in Figure~\ref{fig:path}. The original recordings range from 55~s to 4.5~mn in duration. In~\cite{gloaguen2017}, the extracts are manually annotated and classified in terms of ambiance (\textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}). The available annotations include:

@@

The 75 simulated scenes of the listening test corpus are generated using \textit{simScene} in the \textit{generate} mode with the extracted distributions as input. To improve the diversity of scenarios considered in the listening test, 200 sound scenes are first simulated, equally distributed among the five ambiances (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}). A common duration of 45 seconds is chosen due to the limited duration of listening tests because of the fatigue. Though many studies dealing with the environmental acoustic quality are associated with acoustic measurements ranging from a few seconds~\cite{paulsen1997, brambilla2006} to fifteen minutes~\cite{kuwano1997, decoensel2007} and even to eighty minutes~\cite{namba1988}, stimuli between 30 seconds and 1 minute are often preferred for laboratory tests. To avoid simulated scenes to be too similar in the listening experiment corpus, a selection of 75 simulated scenes is carried out from the initial 200 scenes, in order to cover a large diversity of presence (ranging from no presence to 100\% of presence) for each type of sources without redundancy. In the case of three sources, a 3-dimensional space is constructed where each dimension corresponds to one of the three sources. The selection is then operated by taking the most isolated sound scenes, where the euclidean distance in this space is considered as a measure of the differences between sound scenes.

Playback sound levels are realistically drawn for each scene, and range from 46.6~dB SPL to 77.1~dB SPL over the 75 simulated scenes.

In order to validate the generation procedure of new scenarios, the listening test corpus is extended with replications of recorded sound scenes. Reference recordings are obtained from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations (noted P1-19) with diverse environments. The path undertaken during these soundwalks is shown in Figure~\ref{fig:path}, with sound levels in dB SPL interpolated from measurements at the 19 locations. For each of the 19 corresponding recordings, 45 seconds of audio in a single channel are extracted. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The manual annotations available in~\cite{gloaguen2017} of background and event information are then used to replicate the sound scenes using \textit{simScene} in the \textit{replicate} mode, using the same isolated samples database as for the 75 simulated scenes. The 19 resulting sound scenes are thus simulated, but their scenarios follow those of the reference recordings to the extent of annotation precision. Furthermore, the original 45~s recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization. The 6 recorded and 19 replicated scenes are normalized using the relation found in Section~\ref{sec:exp_equi} so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL over the 19 locations.

@@

During the test the scenes are played at a given sound level as discussed in Section~\ref{sec:exp_stim}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used by all participants. The calibration of the headphones was carried out in a free field situation (pink noise through a Genelec 1031A loudspeaker in front of the head) and consisted in characterizing the relationship between voltage at the headphone terminal and the corresponding binaural sound pressure. To do so, the following procedure was conducted:
\begin{enumerate}
\item A pink noise generator is set at an arbitrary level and its output voltage is measured.
\item Small DPA 4060 microphones are set at the entrance of the ear canals of a human participant~\cite{moller1992}. The generator is used as input of the headphones, placed over the head of the participant. The voltage at the output of the binaural microphones is measured.
\item The headphones are removed, and the generator is used as input of Genelec 1031A loudspeaker placed at a distance of 1 meter from the participant's head. The loudspeaker's amplification is tuned until the same voltage is outputted by the binaural microphones.
\item The head is replaced by a sound level meter to measure the sound level in dB of the loudspeaker. This corresponds to the binaural sound level produced by the headphones for the considered output voltage of the pink noise generator.
\end{enumerate}
By repeating this procedure for different settings of the generator level, the relation between the logarithm of the generator output voltage and headphones playback sound level in dB is approximated as a linear function. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener.

@@

\begin{table*}[t!]
\centering
\caption{Mean differences of perceptual assessments (resp. Pleasantness, Liveliness, Overall Loudness, Interest, Calmness, Time of presence of Traffic, Voices and Birds) between original and replicated sound scenes. Significant differences as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
\label{tab:ogrep}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P1 & 0.43 & \textbf{-1.65} & \textbf{-1.04} & 0.43 & 0.13 & 0.39 & \textbf{-2.09} & 0.61 \\
	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 1.04 & \textbf{-4} & 0.22 \\
	P4 & 0.91 & 0 & \textbf{-1.83} & 0.48 & \textbf{1.30} & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-0.91} & 0.09 & \textbf{-1.43} \\
	P15 & \textbf{-1.35} & 0.52 & 0.52 & \textbf{-1.17} & 0.09 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
	P18 & \textbf{1.13} & -0.30 & \textbf{-1.17} & -0.43 & \textbf{1.39} & \textbf{-1.83} & \textbf{0.83} & \textbf{1.30} \\ \hline
\end{tabular}
%}
\end{table*}

Participants are asked to evaluate sound scenes using 8 perceptual attributes represented by 11-point semantic differential rating scales (0-10). These attributes are presented in French and translated in this paper using standard terminology. The first 5 semantic differential scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, tranquil} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. Additionally, to assess the perceived source activity 3 questions are presented to the participants and evaluated on the same 11-point semantic differential scales:
\begin{itemize}
\item Time of presence of traffic, voice and bird sources: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

@@

The statistic analyses carried out in this paper (statistic comparison tests, principal component analysis) are done with Matlab R2015b (Statistics and Machine Learning Toolbox v10.1). The effect of the \textit{simScene} generation procedure on perception is first investigated. To do so, perceptual responses are compared for the 6 recorded scenes and 6 corresponding replicated scenes, which share common scene compositions. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and semantic differential scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the first five scales, all mean differences are lower than 2 points. Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness: they share a common \textit{voice} class which is characterized in the isolated samples database by recordings of read texts. Though, overall no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.

@@

First, some indicators can be computed directly from the mixed audio. Typically this includes indicators derived from sound level measurements used in monitoring applications. For this study the following variables are considered, and computed with a time frame of 1~s using the Matlab ITA-toolbox~\cite{itatoolbox2017}:
\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is often associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\item $LA_{10}-LA_{90}$: Emergence indicator included in the pleasantness model presented in~\cite{ricciardi2014}.
\end{itemize}
The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of perceived source activity. Its expression is:

\begin{equation}
TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f_1=31.5Hz}^{f_1=16kHz}\lvert\frac{d^2L(f_1, t)}{df_1dt}\rvert}
\end{equation}

where $L(f, t)$ is the third-octave spectrum of the signal. This indicator represents the variations in both the time and frequency dimensions to highlight sources of interest. For example, bird activity is characterized by narrow-band energy with fast paced variations in time, which translates into high $TFSD$ values in the corresponding frequency range. The $TFSD$ is computed for the 4kHz band and 125ms measurements ($TFSD_{4kHz(1/8s)}$), and for the 500Hz band with 1s measurements ($TFSD_{500Hz, 1s}$) as estimates of the perceived activity of birds and voices respectively.

@@

\begin{equation}\label{eq:ts}
\hat T_s(\alpha, \beta) = \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta L_{s}(t, f)\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}>\beta \right]\\
\end{equation}

where $s$ denotes the sound source, $N_t$ is the number of considered time frames, $N_f$ is the number of frequency bands. This indicator evaluates the of percentage of presence of a given source with regard to other sources. First, for each time frame, frequency bands for which the source is emergent, that is the difference between its sound level and that of all other sources $\Delta L_{s}(t, f)$ is greater than the $\alpha$ threshold value, are isolated. Then, the source is considered present in this time frame if the emergence for these bands is on average greater than the $\beta$ threshold. A time of presence estimation is obtained for each source by averaging the resulting binary presence values along the $N_t$ time frames composing the scene. The optimal threshold values are found via grid search so that the time of presence estimation $\hat T_s(\alpha_{opt}, \beta_{opt})$ is best correlated with the average perceived time of presence $T_{s, p}$ as obtained by human evaluation:
\begin{equation}\label{eq:ts_opt}
\alpha_{opt},\beta_{opt} = \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, \hat T_s(\alpha, \beta)\right)
\end{equation}
where $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $r$ denotes the Pearson correlation coefficient and $T_{s, p}$ corresponds to the perceived time of presence assessments averaged per scene. Both thresholds are optimized once in this study using data from the listening test in Section~\ref{sec:exp}, and equal $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$, though several other pairs of values yielded close performance as per the optimization metric defined in eq.\ref{eq:ts}.

@@

\begin{table*}[ht!]
\centering
%\resizebox{\textwidth}{!}{
\caption{Pearson's correlation coefficients between physical and perceptual (resp. Pleasantness, Liveliness, Overall Loudness, Interest, Calmness, Time of presence of Traffic, Voices and Birds) indicators (n = 92).}
\label{tab:physc}
\begin{threeparttable}[t]
\begin{tabular}{ c | c c c c c | c c c c }
\hline
	 & P & L & OL & I & C & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	$LA_{eq}$ & -0.86** & 0.68** & 0.92** & -0.37** & -0.88** & 0.66** & 0.07 & -0.41** \\
	$LA_{50}$ & -0.84** & 0.67** & 0.91** & -0.33** & -0.87** & 0.63** & 0.06 & -0.35** \\
	$L_{eq}$ & -0.88** & 0.67** & 0.91** & -0.44** & -0.88** & 0.71** & 0.06 & -0.46** \\
	$L_{10}$ & -0.87** & 0.65** & 0.90** & -0.44** & -0.86** & 0.71** & 0.06 & -0.47** \\
	$L_{50}$ & -0.89** & 0.65** & 0.92** & -0.43** & -0.89** & 0.71** & 0.03 & -0.44** \\
	$L_{90}$ & -0.86** & 0.68** & 0.92** & -0.39** & -0.89** & 0.67** & 0.07 & -0.40** \\
	$L_{50, 1kHz}$ & -0.88** & 0.69** & 0.92** & -0.42** & -0.89** & 0.73** & 0.08 & -0.50** \\
	$L_{10}-L_{90}$ & 0.13 & -0.18 & -0.24* & -0.06 & -0.22* & -0.01 & -0.01 & -0.09 \\ \hline
	$TFSD_{500Hz, 1s}$ & 0.07 & 0.41** & 0.11 & 0.28** & -0.15 & -0.39** & 0.74** & -0.17 \\
	$TFSD_{4kHz, 1/8s}$ & 0.52** & -0.43** & -0.49** & 0.41** & 0.52** & -0.54** & -0.18 & 0.63** \\ \hline
	$L_{eq, T}$ & -0.58** & 0.20 & 0.46** & -0.46** & -0.42** & 0.71** & -0.16 & -0.36** \\
	$L_{eq, V}$ & -0.17 & 0.50** & 0.31** & 0.08 & -0.37** & -0.04 & 0.71** & -0.40** \\
	$L_{eq, B}$ & 0.27* & -0.04 & -0.11 & 0.35** & 0.18 & -0.24* & -0.04 & 0.71** \\ \hline
	$\Delta L_T$ & -0.45** & -0.11 & 0.26* & -0.59** & -0.22* & 0.66** & -0.51** & -0.26* \\
	$\Delta L_V$ & 0.04 & 0.50** & 0.17 & 0.35** & -0.20 & -0.38** & 0.59** & -0.01 \\
	$\Delta L_B$ & 0.21* & -0.25* & -0.26* & 0.08 & 0.25* & -0.25* & -0.10 & -0.03 \\ \hline
	$\hat T_T(\alpha_{opt}, \beta_{opt})$ & -0.53** & -0.05 & 0.35** & -0.57** & -0.29** & 0.81** & -0.39** & -0.37** \\
	$\hat T_V(\alpha_{opt}, \beta_{opt})$ & 0.12 & 0.44** & 0.05 & 0.35** & -0.11 & -0.39** & 0.81** & -0.16 \\
	$\hat T_B(\alpha_{opt}, \beta_{opt})$ & 0.56** & -0.30** & -0.46** & 0.55** & 0.51** & -0.57** & -0.08 & 0.91** \\ \hline
\end{tabular}
\begin{tablenotes}
\item *: p<0.05, **: p<0.01
\end{tablenotes}
%}
\end{threeparttable}
\end{table*}

Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments, with two-tailed significance tests (see Matlab \textit{corrcoef}). First, all global sound level indicators correlate well (r>0.9) with the perceived overall loudness. Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source $s$ (r=0.71). Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weak for other sources. The proposed estimates of the time of presence $\hat T_s(\alpha_{opt}, \beta_{opt})$ show strong correlations to their perceived counterparts (r>0.8), though this is expected as $\alpha_{opt}$ and $\beta_{opt}$ are optimized to this aim. They also display good source discrimination properties, as no significant correlation is found between voices and birds, and perceptual assessments of traffic were already correlated with those of other sources in Table~\ref{tab:percc}. They are also better predictors than the $TFSD_{500Hz, 1s}$ and $TFSD_{4kHz, 1/8s}$ indicators for voice and bird activity in this corpus. Thus, because source contributions are available in simulated scenes they can be automatically annotated in terms of estimated perceived time of presence using the $\hat T_s(\alpha_{opt}, \beta_{opt})$ indicator for the three sources.
